
# Generating Randomization Sequences {#sequences}

_Contributors: Ben Arnold_


## Introduction

This chapter includes guides for specific types of randomization. It includes code snippets to illustrate the different types of randomization.  In the examples below, we focus on trials with two groups and equal allocation. Multi-arm trials are conceptually similar.  Unequal allocation is similar as well, but raises particular issues around masking, described in the [Masking](#masking) chapter.

In all of the examples, below, we will use masked examples with 2 letter per arm (so 4 letters total).  The rationale for 2+ letters per arm is that if a single letter is accidentally unmasked, then the entire trial will not be unmasked. 

These examples will carry forward into the next chapter on [Randomization Diagnostics](#randomizationdiagnostics). The template files stored in this guide's Github respository (**TBD**) provide complete examples of how to put everything together.

_Note_ In thes examples, we have a list of study IDs that we bind to the allocation sequence. In some studies, such as those that use the REDCap randomization module, we would simply generate a random sequence of treatment assignments that are not bound to specific study IDs.  For example, in REDCap's randomization module it joins study IDs to a randomized treatment assignment (from the generated sequence) within REDCap itself. We simply upload the random sequence to REDCap.

**Essential background reading:**

Altman, D. G. & Martin Bland, J. Treatment allocation in controlled trials: why randomise? BMJ 318, 1209–1209 (1999). https://pubmed.ncbi.nlm.nih.gov/10221955/ [@Altman1999-zc]

Altman, D. G. & Bland, J. M. How to randomise. BMJ 319, 703–704 (1999). https://pubmed.ncbi.nlm.nih.gov/10480833/ [@Altman1999-da]

Schulz, K. F. & Grimes, D. A. Generation of allocation sequences in randomised trials: chance, not choice. Lancet 359, 515–519 (2002). https://pubmed.ncbi.nlm.nih.gov/11853818/ [@Schulz2002-jf]


## Simple randomization

> _"No other allocation generation approach, irrespective of its complexity and sophistication, surpasses the unpredictability and bias prevention of simple randomisation."_ [@Schulz2002-jf] 

High praise from masters of clinical trials! Simple (unrestricted) randomization is the purest form of randomization. Patients (or clusters) are allocated to each arm with a known (usually equal) probability. For larger trials (hundreds of units) without specific design considerations this might be just the right tool. Many trials led by Proctor faculty favor simple randomization. 

Schulz and Grimes highlight a downside, which is relevant for smaller trials or those with stratification into smaller groups.  With simple randomization, at any point in the sequence the numbers of patients allocated to each treatment group will probably differ, which is not a major problem in terms of statistical efficiency (though perfectly balanced is statistically ideal).  For a two-arm trial, the chance of pronounced imbalance becomes negligible with trial sizes greater than 200 [@Lachin1988-gg; @Schulz2002-jf]. But, if 20 participants are randomized with equal probability to two treatment groups, the chance of a 12:8 split (i.e., 60% A, 40% B) or worse is approximately 50%. For 100 participants, the chance of the same ratio (60:40 split) or worse is only 5% [@Friedman2015-ri]. For smaller trials and multi-center trials this chance imbalance could be a problem, and so many trials use blocking and/or stratification in the randomization sequence (next sections).

Additional advice from Schulz and Grimes [@Schulz2002-xo]:

> _"Overall, investigators underuse simple randomisation and overuse fixed-block randomisation. For non-double-blinded trials larger than 200 participants, investigators should use simple randomisation more often and accept moderate disparities in group sizes."_

Here is an example of generating a simple randomization sequence: 

```{r simple randomization sequence generation}
#-------------------------------
# Example of simple randomization
#
# 2 arm trial
# 2 letters per arm (masked)
#-------------------------------

#-------------------------------
# create a list of 1000 IDs.
# in a real trial, we may already
# have a list or sampling frame 
# to work with 
#
# pro tip: always ensure your IDs start with a character
# to prevent accidental conversion to numeric variables!
#-------------------------------
nobs <- 1000
id <- paste0("ID",1:nobs)

#-------------------------------
# Identify letters for
# the randomization
#
# In a masked trial, we would
# start with placeholder letters
# to vet the sequence generation
# and then change them for the
# final randomization.
# See step 3 in the Overview
#
# In this example, we use
# A, B, T, V
#-------------------------------
rand_letters <- c("A","B","T","V")

#-------------------------------
# set a seed
# 
# this is essential for 
# creating a reproducible 
# sequence!  After vetting the
# randomization with the broader
# study team (steps 4-5 in the Overview), 
# we can change the seed to create a new
# (protected) sequence
#-------------------------------
set.seed(123)

#-------------------------------
# generate the sequence
#
# the sample() function
# samples the rand_letters
# with equal probability
# up to the number of IDs
# using replacement
#-------------------------------
tr_masked <- sample(rand_letters, size = length(id), replace = TRUE)

#-------------------------------
# combine the IDs with the
# sequence in a data.frame
#
# this object would be stored
# in the Randomization Directory
# as a .csv file
#-------------------------------
rand_seq <- data.frame(id = id, tr_masked = tr_masked)


#-------------------------------
# for the sake of example,
# unmask and examine the 
# allocation. 
#
# Pretend letters A and T 
# are for placebo
#-------------------------------
tr <- ifelse(tr_masked %in% c("A","T"),"Placebo","Treatment")
table(tr)

```

In this example, 51% of the participants were allocated to placebom and 49% were allocated to treatment  — a pretty good balance! This would be expected given the trial's large size.  

Try running the code above with `nobs <- 80` to see what a typical randomization looks like:

```{r simple randomization sequence generation with small sample size}
#-------------------------------
# From the above example
# create a function
# that takes as arguments
# a vector of IDs and
# masked treatment letters
#-------------------------------
randomize_simple <- function(ids, letters) {
  tr_masked <- sample(rand_letters, size = length(id), replace = TRUE)
  return( data.frame(id = ids, tr_masked = tr_masked) )
}

#-------------------------------
# setting trial size, creating IDs
# and identifying letters,
# as in the above example
#-------------------------------
nobs <- 80
id <- paste0("ID",1:nobs)
rand_letters <- c("A","B","T","V")
#-------------------------------
# generating a new sequence
# using the randomize_simple()
# function
#-------------------------------
set.seed(123)
rand_seq2 <- randomize_simple(ids = id, letters = rand_letters)
rand_seq2$tr <- ifelse(rand_seq2$tr_masked %in% c("A","T"),"Placebo","Treatment")
table(rand_seq2$tr)

```

With the same seed, 48/80 (60%) are allocated to placebo, which is not ideal.  

Blocking can prevent this type of chance imbalance.

## Blocked randomization

> _"The method of restricted randomisation is used to balance sample sizes. That balance usually enhances statistical power and addresses any time trends that might exist in treatment efficacy and outcome measurement during the course of a trial. Moreover, restricted randomisation within strata becomes essential for investigators to attain the benefits of stratification. Thus, reasonable scientific justification lends support to restriction."_ [@Schulz2002-xo]

Blocking is one form of restricted randomization. It is often combined with other forms of restriction, including stratification and matching (next sections).  

With blocking, the numbers of patients randomized in the two groups can never differ by more than half the block length. If the trial uses multiple letters to mask groups, then the number of patients in the two groups can never differ by more than half the block length divided by the number of letters per group.

A simple randomization sequence will translate into a random string of letters, such as `A A B B B B A B` for the first eight patients.  If we consider blocks of four at a time, there are only 6 ways in which two patients get `A` and two patients get `B`: 1: `A A B B` 2: `A B A B` 3: `A B B A` 4: `B B A A` 5: `B A B A` 6: `B A A B`.  

Since this could create a predictable sequence within blocks, we recommend the use of randomly permuted block lengths [@Schulz2002-xo].  For example, using block sizes of 2, 4 and 6, each with a probability.  Note that the block length needs to be a multiple of the number of letters used in the randomization scheme and the sum of the block size probabilities must sum to 1.  Blocks that are too long do not preserve balance as well.  For a two arm trial, we often use block sizes of 4 and 6.  If multiple letters are used per group, then we can use slightly longer block lengths, as in the example below.

> It is best practice to keep the block sizes and block probabilities secret to help with allocation concealment [@Schulz1995-lb].

Here is an example of generating a blocked randomization sequence. Drawing from the basics in the example above, the sequence generation uses permuted blocks of length 8 and 12 (multiples of the 4 masked letters used). Since there are 2 letters per group,  the maximum sequence length divided by number of letters per group (12/2) = 6, so the most the sequence could ever be imbalanced is 1/2*6 = 3

```{r blocked randomization sequence generation}
#-------------------------------
# setting trial size, creating IDs
# and identifying letters,
# as in the above example
#----------------------------
nobs <- 80
id <- paste0("ID",1:nobs)
rand_letters <- c("A","B","T","V")

#-------------------------------
# set a seed for reproducibility!
#-------------------------------
set.seed(233)

#-------------------------------
# generate the permuted block
# sequence
# block lengths of 8, 12
# with probabilities 0.7, 0.3
#-------------------------------
# the number of permuted blocks can be long, much longer than needed
nblocks <- 100
block_len <- sample(c(8,12),size=nblocks,replace=TRUE,prob=c(0.7,0.3))

#-------------------------------
# expand the sequence, keep
# track of the block lengths
#
# (note: in this step and the next
#  there is a more transparent
#  way to code this using dplyr/tidyr,
#  but we are trying to rely,
#  when possible, on base R for stability)
#-------------------------------
d_seq <- data.frame( 
  block_id = rep(seq_len(nblocks), block_len), 
  block_size = rep(block_len, block_len)
  )

#-------------------------------
# generate the allocation sequence
#
# equal allocation within 
# permuted blocks of size 8 or 12
#
# the loop below iterates through
# each block generated above,
# and randomly samples the masked
# letters (rand_letters) for 
# either 8 or 12, depending on the
# randomly permuted block length
#-------------------------------
d_seq$tr_masked <- as.character(NA)
for(blocki in unique(d_seq$block_id) ) {
  d_seq$tr_masked[d_seq$block_id == blocki] <- ifelse(
    d_seq$block_size[d_seq$block_id == blocki] == 8, 
    sample(rep(rand_letters,2), size=8,  replace=FALSE),
    sample(rep(rand_letters,3), size=12, replace=FALSE)
    )
}

#-------------------------------
# for the sake of example,
# unmask and examine the 
# allocation for the first nobs
#
# Pretend letters A and T 
# are for placebo
#-------------------------------
d_seq$tr <- ifelse(d_seq$tr_masked %in% c("A","T"),"Placebo","Treatment")
table(d_seq$tr[1:nobs])

```

In this example, the treatment sequence is now very well balanced with a small sample size of just 80 units.  In the next chapter on [Randomization Diagnostics](#randomizationdiagnostics), we will illustrate how to examine balance across arms over cumulative enrollment as an internal validity check to ensure the sequence is consistent with the blocking scheme.

## Stratified randomization


## Matched randomization





